### 2016-03-14 HSF-TN/draft-2016-PKG/draft-HSF-TN-2016-PKG.tex
PKG 2015->2016

\\
 & & HSF-TN-2015-XXX \\  % ID 
 & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

@@

This document describes the discussions and findings of the working group.
\subsection{The Working Group}
The working group consisted of members of the HEP community from the energy and intensity frontiers as well as accelerator modeling. The group met in a series of meetings over the months of May and June~\cite{meetings}.
Additional discussions took place via the public \href{https://groups.google.com/forum/#!forum/hep-sf-packaging-wg}{packaging working group mailing list}.
\subsection{The Packaging Problem Space}
Collectively across HEP and even within some large institutions there is
a need to maintain software installations which span a broad
configuration space.  It is helpful to consider this as an actual
multi-dimensional space and decompose it along the following orthogonal
and discrete dimensions:

@@

\item[variant] describes the set of options applied during the process
of building a particular (project,version) pair.  The variant of a
package is in general itself multi-dimensional.  Between any given set
of (package,version) points there is some amount of degeneracy in the
variant dimension.  For example, it is common now that all C++ packages
that must work together must be built with a consistent compiler and C++
standard.  However, some packages may be built with optimization and no
debug symbols and others, vice versa.

@@

A single (\textit{project,version,variant}) point in this space
represents one built \textit{package} (for now, ignoring details of file
system layout, bundling, distribution, run-time environment).  The
software that is provided by the operating system may be thought of as a
single point in this space.

@@

This picture allows the expression of some required and desired featues
of a packaging system:
% these remain to be discussed, for sure!

@@

\item require ability to derive a novel package-space tree from an
existing one (eg, upgrade ROOT to new version, make debug/opt versions
of some packages, use native compiler or build one as part of the tree).

\end{itemize}
Many HEP experiments define their own naming conventions for the \textit{variant} dimension in this space, and they are all equally valid but inhibit sharing due to trivial language differences.  A sub-goal of this working group is to agree on a field wide naming convention. The HSF Packaging group has a draft proposal, available here: https://github.com/HEP-SF/documents/blob/master/HSF-TN/draft-2015-NAM/draft-HSF-TN-2015-NAM.pdf
% OK there is probably a better way to reference this but I don't know it... -LSK

@@

Software stacks are needed by many organizations most notably HEP experiments.  If a package is authored by the organization it is not considered an external package however it still will need build recipes. Any system that builds external packages into a coherent whole must also be capable of building the internal packages as well.  In the area of external packages it will be useful to distinguish between ``foreign'' and ``domain'' external packages. Domain packages are authored by groups within the field of HEP and therefor the HSF could influence the behavior of the domain packages. The foreign packages have to be dealt with as they exist. Examples of foreign packages include: gcc, fftw3, eigen and lapack. Examples of domain packages include clhep and ROOT. In the following a few of the solutions used within the community are summarized. For more details we refer to the project websites.

@@

\emph{LCGCMake}\cite{lcgcmake} is the build tool used for creating the LCG releases used by the LHC experiments ATLAS and LHCb. It is based on a set of CMake macros and each external package is described using one of these macros in a semi-declarative style.

@@

\subsection{Solution 6 -- aliBuild}
\emph{aliBuild}\cite{aliBuild} is an evolution of cmsBuild that fixes many of the weakness of cmsBuild. Relative to the above it is not dependent on rpm and builds tar-files that can be currated into a number of diffent package managers such as yum, and apt.
% I need to get Giulio to sign off on my short description. -LSK

@@

\subsection{Solution 9 - Spack }
\emph{Spack}\cite{Spack} is a packaging system specifically developed for the HPC use case which has a lot in common with the HEP requirements.
``Spack provides a novel, re-cursive specification syntax to invoke parametric builds of packages and dependencies. It allows any number of builds to coexist on the same system, and it ensures that installed packages can find their dependencies, regardless of the environment." On creating Spack the authors did a review of existing packaging systems, including a few of those discussed in this document.

@@

\subsection{Supported Platforms and Environments}
The most basic criterion is the support for the various platforms and environments used in high-energy physics. There are three categories of environments:
\begin{enumerate}
\item \textbf{Linux}
\item \textbf{MacOS X}
\item \textbf{Windows}
\end{enumerate}
The first category is split into several flavours of distributions. The main distribution are the RedHat derived Scientific Linux~\cite{sl} and CentOS~\cite{centos}. Compared to these, Debian~\cite{debian} and Debian based based distributions play a smaller role in computing centres. On desktops, the Debian based Ubuntu~\cite{ubuntu} seems rather popular.

@@

\subsection{Build and Installation Variants}
One important feature of build tools in HEP is the support for the installation of multiple stacks or package versions in parallel. 
\begin{enumerate}
\item \textbf{Multi-Rel}: The support for building and installing multiple stacks in parallel. This is important if multiple versions of experiment software need to coexist on a given installation.
\item \textbf{Multi-BuildVar}: The support for multiple varients of the same package, release pair within the same stack. There are many possible reasons why this maybe needed, what they have in common is the need to specify compiler switches that have to be applied consistently across the build, like a debugging option or a compiler dialect.
\item \textbf{MultiShell-RTE}: The support for setting up the runtime environment for a given stack triplet (project,version,variant), supporting multiple shell flavours.
\item \textbf{Relocation}: Whether the build tool supports the relocation of packages or creates fully relocatable packages.
\end{enumerate}

@@

\begin{table}
\centering
\begin{tabular}{lcccc}
&Multi-Rel & Multi-BuildVar & MultiShell-RTE & Relocation \\ \hline
aliBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} & \cellcolor{green} + \\
cmsBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} & \cellcolor{green} + \\
Contractor & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} NA  &  \cellcolor{red} - \\
Homebrew &  \cellcolor{red} - & \cellcolor{red} -   & \cellcolor{yellow} NA   & \cellcolor{red} -  \\
LCGCMake & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} +\\
Nix & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} +  & \cellcolor{red} -\\
SciSoft &  \cellcolor{green} + & \cellcolor{green} + &  \cellcolor{green} + & \cellcolor{green} + \\
Spack & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} NA & \cellcolor{red} -\\
Worch &  \cellcolor{green} + &  \cellcolor{green}  + &  \cellcolor{green} + &  \cellcolor{green} +
\end{tabular}
\caption{\label{tab:taxonomyVariants}Support for multiple releases and build variants. \emph{NA} denotes that the criteria are not-applicable for homebrew and Spack. As it only provides one variant it can rely on default system paths.}
\end{table}

@@

The findings are summarized in Table~\ref{tab:taxonomyUsability}. The dependencies of most of the tools seem well under control. The addition of new packages is of similar complexity in all cases. Only the ease of bootstrapping differs. For most of them it is a simple checkout. However, others like Nix an entire (fake-)root environment is necessary. The documentation of the domain specific packages is in general very poor, the documentation of the non-HEP tools surprisingly good and complete.

@@

\subsection{Other Criteria}
During the meetings proponents brought up other creiteria that are not so easy to group together.  Some might consider these requirements, other might disagree. For now we've just listed them as ``other''. The criteria in this category are:
\begin{enumerate}
\item \textbf{Performance}: The relevant metrics are a. length of time to build the entire stack, this includes support for parallel builds. b. length of time to incrementally build a developer defined subset of the stack. c. Ability to reuse binary packages to speedup builds on different machines.
%BH: I merged Parallel + Performance again
\item \textbf{Sys-Reuse}: the ability of the build system to reuse parts of the system software that is being built if desired.
\item \textbf{Community}: whether the tool is used by a wider community or not. Green means it is a standard opensource project (e.g. homebrew, conda, nix), yellow means it is HEP / science only community, red means it is a single experiment or person effort.
\item \textbf{Unique-IDs}: A method of uniquely identifying a build product such that if it already exists, it does not have to be built again. This feature insures that the software suite will always be built consistently throughout the whole stack. Further discussion with the group reveals that this point is complicated. We will expand on this in the next section.
%It is also desirable to have the ability to develop a local copy of the package and still build using the tool (including noticing changes in the local area and rebuilding a consistent build).
  \item \textbf{VCS-Support} Integrated support for check-out of given tags or branches from the projects’ repositories directly.
\end{enumerate}
The findings are summarized in Table~\ref{tab:taxonomyOthers}. 
\begin{table}
\centering
\begin{tabular}{lcccccc}

@@

\begin{thebibliography}{9}
\bibitem{meetings} Indico agendas of \href{https://indico.cern.ch/event/373973/}{25.2.2015}, \href{https://indico.cern.ch/event/398344/}{2.6.2015}, \href{https://indico.cern.ch/event/400272/}{9.6.2015}, \href{https://indico.cern.ch/event/402229/}{16.6.2015}, \href{https://indico.cern.ch/event/403790/}{23.6.2015}

\bibitem{worch} Worch: \url{https://github.com/brettviren/worch}
\bibitem{lcgcmake} LCGCMake: \url{http://ph-dep-sft.web.cern.ch/document/using-lcgcmake}
\bibitem{cmsbuild} cmsBuild: \url{https://github.com/cmsbuild/cmsdist}
\bibitem{aliBuild} aliBuild: \url{http://alisw.github.io/alibuild/}
\bibitem{Spack} Spack: \url{http://www.computer.org/csdl/proceedings/sc/2015/3723/00/2807623.pdf}

\bibitem{contractor} Contractor: \url{https://cdcvs.fnal.gov/redmine/projects/contractor/wiki}
\bibitem{homebrew} Homebrew: \url{http://brew.sh/}
\bibitem{nix} Nix: \url{https://nixos.org/nix/}

>>>>

\\
 & & HSF-TN-2016-XXX \\  % ID
 & & \today \\ % Date - Can also hardwire e.g.: 23 March 2010
 & & \\
% not in paper \hline
\end{tabular*}

@@

This document describes the discussions and findings of the working group.
\subsection{The Working Group}
The working group consisted of members of the HEP community from the energy and intensity frontiers as well as accelerator modeling. The group met in a series of meetings over the months of May and June 2015~\cite{meetings}.
Additional discussions took place via the public \href{https://groups.google.com/forum/#!forum/hep-sf-packaging-wg}{packaging working group mailing list}.
\subsection{The Packaging Problem Space}
Collectively across HEP and even within some large institutions there is
a need to maintain software installations which span a broad
configuration space.  It is helpful to consider this as an actual
multi-dimensional space and decompose it along the following orthogonal
and discrete dimensions:

@@

\item[variant] describes the set of options applied during the process
of building a particular (project,version) pair.  The variant of a
package is in general itself multi-dimensional.  Between any given set
of (package,version) points there is some amount of degeneracy in the
variant dimension.  For example, it is common now that all C++ packages
that must work together must be built with a consistent compiler and C++
standard.  However, some packages may be built with optimization and no
debug symbols and others, vice versa. Debug symbols and the choice of a
particular C++ standard are the most used variants in HEP.

\item[platform] denotes the combination of operating system, hardware
architecture and chosen compiler and compiler version. Many HEP experiments
define their own naming conventions for the \textit{platform} dimension in this
space, and they are all equally valid but inhibit sharing due to trivial
language differences.  A sub-goal of this working group is to agree on a field
wide naming convention, in form of another technical
note~\cite{naming-conventions}.

\item[dependencies] of a particular build of a project can be of different
version and variant. The chosen setup of the dependencies are thus
another dimension in the configuration space.

@@

For the purpose of this note a single (\textit{project,version,variant,platform,dependencies})
point in this space is called a built \textit{package} (for now, ignoring details of file
system layout, bundling, distribution, run-time environment).  The
software that is provided by the operating system may be thought of as a
single point in this space.

@@

This picture allows the expression of some required and desired features
of a packaging system:
% these remain to be discussed, for sure!

@@

\item require ability to derive a novel package-space tree from an
existing one (eg, upgrade ROOT to new version, make debug/opt versions
of some projects, use native compiler or build one as part of the tree).

\end{itemize}

@@

Software stacks are needed by many organizations most notably HEP experiments.
If a project is authored by the organization it is not considered an external
project however it still will need build recipes. Any system that builds
external projects into a coherent whole must also be capable of building the
internal projects as well.  In the area of external projects it will be useful
to distinguish between ``foreign'' and ``domain'' external projects. Domain
projects are authored by groups within the field of HEP and therefor the HSF
could influence the behavior of the domain projects. The foreign projects have
to be dealt with as they exist. Examples of foreign projects include:
gcc, fftw3, eigen and lapack. Examples of domain projects include clhep and
ROOT. In the following a few of the solutions used within the community are
summarized. For more details we refer to the project websites.

@@

\emph{LCGCMake}\cite{lcgcmake} is the build tool used for creating the LCG
releases used by the LHC experiments ATLAS and LHCb. It is based on a set of
CMake macros and each external project is described using one of these macros
in a semi-declarative style.

@@

\subsection{Solution 6 -- aliBuild}
\emph{aliBuild}\cite{aliBuild} is an evolution of cmsBuild that fixes many of the weakness of cmsBuild. Relative to the above it is not dependent on rpm and builds tar-files that can be curated into a number of diffent package managers such as yum, and apt.
% I need to get Giulio to sign off on my short description. -LSK

@@

\subsection{Solution 9 - Spack }
\emph{Spack}\cite{Spack} is a packaging system specifically developed for the HPC use case which has a lot in common with the HEP requirements.
``Spack provides a novel, recursive specification syntax to invoke parametric builds of packages and dependencies. It allows any number of builds to coexist on the same system, and it ensures that installed packages can find their dependencies, regardless of the environment." On creating Spack the authors did a review of existing packaging systems, including a few of those discussed in this document.

@@

\subsection{Supported Platforms and Environments}
The most basic criterion is the support for the various platforms and
environments used in high-energy physics. There are three categories of
environments:
\begin{enumerate}
\item \textbf{Linux}
\item \textbf{MacOS X}
\item \textbf{Windows}
\end{enumerate}
The first category is split into several flavours of distributions. The main
distribution are the RedHat derived Scientific Linux~\cite{sl} and
CentOS~\cite{centos}. Compared to these, Debian~\cite{debian} and Debian based
distributions play a smaller role in computing centres. On desktops, the Debian
based Ubuntu~\cite{ubuntu} seems rather popular.

@@

\subsection{Build and Installation Variants}
One important feature of build tools in HEP is the support for the installation of multiple stacks or package versions in parallel.
\begin{enumerate}
\item \textbf{Multi-Rel}: The support for building and installing multiple stacks in parallel. This is important if multiple versions of experiment software need to coexist on a given installation.
\item \textbf{Multi-BuildVar}: The support for multiple variants of the same project, release pair within the same stack. There are many possible reasons why this may be needed, what they have in common is the need to specify compiler switches that have to be applied consistently across the build, like a debugging option or a compiler dialect.
\item \textbf{MultiShell-RTE}: The support for setting up the runtime environment for a given stack triplet (project,version,variant), supporting multiple shell flavours.
\item \textbf{Relocation}: Whether the build tool supports the relocation of packages or creates fully relocatable packages.
\end{enumerate}

@@

\begin{table}
\centering
\begin{tabular}{lcccc}
&Multi-Rel & Multi-BuildVar & MultiShell-RTE & Relocation \\ \hline
aliBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + \\
cmsBuild & \cellcolor{green} +  & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + \\
Contractor & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} NA  &  \cellcolor{red} - \\
Homebrew &  \cellcolor{red} - & \cellcolor{red} -   & \cellcolor{yellow} NA   & \cellcolor{red} -  \\
LCGCMake & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{green} +\\
Nix & \cellcolor{green} + & \cellcolor{red} - & \cellcolor{green} +  & \cellcolor{red} -\\
SciSoft &  \cellcolor{green} + & \cellcolor{green} + &  \cellcolor{green} + & \cellcolor{green} + \\
Spack & \cellcolor{green} + & \cellcolor{green} + & \cellcolor{yellow} NA & \cellcolor{red} -\\
Worch &  \cellcolor{green} + &  \cellcolor{green}  + &  \cellcolor{green} + &  \cellcolor{green} +
\end{tabular}
\caption{\label{tab:taxonomyVariants}Support for multiple releases and build variants. \emph{NA} denotes that the criteria are not-applicable for homebrew and Spack. As it only provides one variant it can rely on default system paths.}
\end{table}

@@

The findings are summarized in Table~\ref{tab:taxonomyUsability}. The
dependencies of most of the tools seem well under control. The addition of new
packages is of similar complexity in all cases. Only the ease of bootstrapping
differs. For most of them it is a simple checkout. However, for others like Nix
an entire (fake-)root environment is necessary. The documentation of the domain
specific projects is in general very poor, the documentation of the non-HEP
tools surprisingly good and complete.

@@

\subsection{Other Criteria}
During the meetings proponents brought up other criteria that are not so easy to group together.  Some might consider these requirements, others might disagree. For now we've just listed them as ``other''. The criteria in this category are:
\begin{enumerate}
\item \textbf{Performance}: The relevant metrics are a. Length of time to build
the entire stack, this includes support for parallel builds. b. Length of time
to incrementally build a developer defined subset of the stack. c. Ability to
reuse binary packages to speedup builds on different machines.
\item \textbf{Sys-Reuse}: the ability of the build system to reuse parts of the system software that is being built if desired.
\item \textbf{Community}: whether the tool is used by a wider community or not. Green means it is a standard opensource project (e.g. homebrew, conda, nix), yellow means it is HEP / science only community, red means it is a single experiment or person effort.
\item \textbf{Unique-IDs}: A method of uniquely identifying a build product such that if it already exists, it does not have to be built again. This feature insures that the software suite will always be built consistently throughout the whole stack. Further discussion with the group reveals that this point is complicated. We will expand on this in the next section.
%It is also desirable to have the ability to develop a local copy of the package and still build using the tool (including noticing changes in the local area and rebuilding a consistent build).
  \item \textbf{VCS-Support} Integrated support for check-out of given tags or branches from the projects’ repositories directly.
\end{enumerate}
The findings are summarized in Table~\ref{tab:taxonomyOthers}.
\begin{table}
\centering
\begin{tabular}{lcccccc}

@@

\begin{thebibliography}{9}
\bibitem{meetings} Indico agendas of
\href{https://indico.cern.ch/event/373973/}{25.2.2015},
\href{https://indico.cern.ch/event/398344/}{2.6.2015},
\href{https://indico.cern.ch/event/400272/}{9.6.2015},
\href{https://indico.cern.ch/event/402229/}{16.6.2015},
\href{https://indico.cern.ch/event/403790/}{23.6.2015},
\href{https://indico.cern.ch/event/457365/}{4.11.2015},
\href{https://indico.cern.ch/event/462334/}{18.11.2015},
\href{https://indico.cern.ch/event/484006/}{10.2.2016}
\bibitem{naming-conventions}Naming convention TN listed on \url{http://hepsoftwarefoundation.org/technical_notes.html}
\bibitem{worch} Worch: \url{https://github.com/brettviren/worch}
\bibitem{lcgcmake} LCGCMake: \url{http://ph-dep-sft.web.cern.ch/document/using-lcgcmake}
\bibitem{cmsbuild} cmsBuild: \url{https://github.com/cmsbuild/cmsdist}
\bibitem{contractor} Contractor: \url{https://cdcvs.fnal.gov/redmine/projects/contractor/wiki}
\bibitem{aliBuild} aliBuild: \url{http://alisw.github.io/alibuild/}
\bibitem{homebrew} Homebrew: \url{http://brew.sh/}
\bibitem{nix} Nix: \url{https://nixos.org/nix/}
\bibitem{Spack} Spack: \url{http://www.computer.org/csdl/proceedings/sc/2015/3723/00/2807623.pdf}
